draft

### 2025-05-26 draft.tex

We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
We then traind a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Panels A and B.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Panel C.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions. \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics. \textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings. \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown). The top left quarter of the image shows that the morphological approach worked to predict regions while the expression annotations did not.}
\end{figure}

@@

% Already, CellProfiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
While developed for biology, engineered morphological features have proven valuable far beyond their original domain—from analyzing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING. 
cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_cpmeasure-3FBF}. 
%The datasets we produced for this work are available on Zenodo. The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}. The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
% 
\subsubsection{Acknowledgements}
\label{sec:acknowledgements}

>>>>

We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Panels A and B.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Panel C.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features, such as the minor axis length, refer to the nucleus, while intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions. \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics. \textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings. \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown). The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
\end{figure}

@@

While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING and agricultural automation \citep{xuDesignNonDestructiveSeed2024}. 
cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}

\onecolumn
\section{Appendix}

\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_cpmeasure-3FBF}. 
%The datasets we produced for this work are available on Zenodo. The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}. The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
%

\subsubsection{Details on the spatial use case}
\label{sec:spatial_extra}

\subsubsection{Acknowledgements}
\label{sec:acknowledgements}
